---
title: "FLARE Data Assimilation Diagnostics"
subtitle: "Filter health, calibration, and hyperparameter guidance"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-tools: true
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 5
execute:
  warning: false
  message: false
params:
  diagnostics_dir: "da_diagnostics"   # path to folder written by write_da_diagnostics()
  site_id: null                        # optional: filter to one site_id
  model_id: null                       # optional: filter to one model_id
---

```{r setup}
library(arrow)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(lubridate)
library(knitr)
library(gt)

theme_set(theme_bw(base_size = 12))

DIAG_DIR <- params$diagnostics_dir

.read_parquet_safe <- function(fname) {
  path <- file.path(DIAG_DIR, fname)
  if (!file.exists(path)) return(NULL)
  df <- arrow::read_parquet(path)
  if (!is.null(params$site_id))  df <- filter(df, site_id  == params$site_id)
  if (!is.null(params$model_id)) df <- filter(df, model_id == params$model_id)
  df
}

fh  <- .read_parquet_safe("da_filter_health.parquet")
inn <- .read_parquet_safe("da_innovations.parquet")
spr <- .read_parquet_safe("da_ensemble_spread.parquet")
rnk <- .read_parquet_safe("da_ranks.parquet")
par <- .read_parquet_safe("da_parameters.parquet")
cst <- .read_parquet_safe("da_cross_state.parquet")
cdp <- .read_parquet_safe("da_cross_depth.parquet")

has_pars      <- !is.null(par) && nrow(par) > 0
has_cross     <- !is.null(cst) && nrow(cst) > 0
n_da_steps <- if (!is.null(fh)) nrow(fh) else 0L
```

```{r meta-banner}
#| results: asis
if (n_da_steps == 0) {
  cat("::: {.callout-warning}\n")
  cat("## No diagnostic data found\n")
  cat("No parquet files were found in `", DIAG_DIR, "`.\n")
  cat("Set `save_da_diagnostics: TRUE` in `configure_flare.yml` and re-run.\n")
  cat(":::\n")
  knitr::knit_exit()
}

site_label  <- if (!is.null(fh$site_id[1]))  fh$site_id[1]  else "unknown"
model_label <- if (!is.null(fh$model_id[1])) fh$model_id[1] else "unknown"
cat(sprintf(
  "**Site:** %s &nbsp; | &nbsp; **Model:** %s &nbsp; | &nbsp; **DA steps:** %d &nbsp; | &nbsp; **Date range:** %s – %s\n\n",
  site_label, model_label, n_da_steps,
  format(min(fh$time), "%Y-%m-%d %H:%M"),
  format(max(fh$time), "%Y-%m-%d %H:%M")
))
```

---

## 1 Filter Health {#sec-filter-health}

> **NIS** (Normalized Innovation Squared) measures whether ensemble spread matches
> forecast error. A well-tuned filter has NIS ≈ 1. **DFS** (Degrees of Freedom for
> Signal) measures how much information is extracted from observations each step.

```{r filter-health-plot}
nis_healthy <- c(0.5, 2.0)

p_nis <- fh |>
  mutate(status = case_when(
    nis < 0.2 | nis > 5   ~ "critical",
    nis < nis_healthy[1] | nis > nis_healthy[2] ~ "warning",
    TRUE                   ~ "healthy"
  )) |>
  ggplot(aes(x = time, y = nis, colour = status)) +
  geom_hline(yintercept = 1,              linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = nis_healthy,    linetype = "dotted", colour = "#d9a520", linewidth = 0.6) +
  geom_hline(yintercept = c(0.2, 5),      linetype = "dotted", colour = "#c0392b", linewidth = 0.6) +
  geom_point(size = 2) +
  geom_line(aes(group = 1), colour = "grey70", linewidth = 0.4) +
  scale_colour_manual(
    values = c(healthy = "#27ae60", warning = "#d9a520", critical = "#c0392b"),
    guide = guide_legend(title = NULL)
  ) +
  scale_y_log10() +
  labs(title = "Normalized Innovation Squared (NIS)", y = "NIS (log scale)", x = NULL) +
  annotate("text", x = min(fh$time), y = 1.05, label = "ideal = 1", hjust = 0,
           size = 3, colour = "grey40")

# Bar width in seconds derived from actual median spacing between DA steps.
# geom_col with POSIXct x always needs width in seconds; as.numeric(difftime)
# can return days/hours depending on scale and produces invisible bars.
bar_width_secs <- if (n_da_steps > 1) {
  as.numeric(median(diff(fh$time)), units = "secs") * 0.85
} else {
  86400 * 0.85
}

p_dfs <- ggplot(fh, aes(x = time, y = dfs)) +
  geom_col(fill = "#2980b9", alpha = 0.7, width = bar_width_secs) +
  labs(title = "Degrees of Freedom for Signal (DFS)", y = "DFS", x = NULL)

p_nobs <- ggplot(fh, aes(x = time, y = n_obs)) +
  geom_col(fill = "#7f8c8d", alpha = 0.7, width = bar_width_secs) +
  labs(title = "Number of Observations Assimilated", y = "n obs", x = NULL)

p_nis / p_dfs / p_nobs + plot_layout(heights = c(2, 1, 1))
```

```{r filter-health-table}
fh |>
  summarise(
    `Mean NIS`   = round(mean(nis,   na.rm = TRUE), 3),
    `Median NIS` = round(median(nis, na.rm = TRUE), 3),
    `% steps NIS > 2` = round(mean(nis > 2,  na.rm = TRUE) * 100, 1),
    `% steps NIS < 0.5` = round(mean(nis < 0.5, na.rm = TRUE) * 100, 1),
    `Mean DFS`   = round(mean(dfs,   na.rm = TRUE), 2),
    `Mean n_obs` = round(mean(n_obs, na.rm = TRUE), 1)
  ) |>
  gt() |>
  tab_header(title = "Filter Health Summary") |>
  tab_style(
    style = cell_fill(color = "#fef9c3"),
    locations = cells_body(columns = everything())
  )
```

---

## 2 Innovation Analysis {#sec-innovations}

> The **innovation** is the difference between what was observed and what the ensemble
> predicted (obs − H·x̄). Persistent non-zero mean innovations for one variable
> indicate systematic model bias. Large spread suggests the observation is surprising
> the ensemble repeatedly.

```{r innovation-summary}
if (!is.null(inn)) {
  inn_summary <- inn |>
    group_by(variable) |>
    summarise(
      n_obs       = n(),
      mean_innov  = round(mean(innovation, na.rm = TRUE), 4),
      sd_innov    = round(sd(innovation,   na.rm = TRUE), 4),
      mean_nis    = round(mean(nis_single, na.rm = TRUE), 3),
      pct_nis_gt2 = round(mean(nis_single > 2, na.rm = TRUE) * 100, 1),
      mean_influence = round(mean(obs_influence, na.rm = TRUE), 3),
      .groups = "drop"
    ) |>
    arrange(desc(mean_nis))

  inn_summary |>
    gt() |>
    tab_header(title = "Innovation Statistics by Observation Type") |>
    cols_label(
      variable = "Obs type", n_obs = "N",
      mean_innov = "Mean innov", sd_innov = "SD innov",
      mean_nis = "Mean NIS", pct_nis_gt2 = "% NIS > 2",
      mean_influence = "Mean influence"
    ) |>
    data_color(
      columns = mean_nis,
      method  = "numeric",
      palette = c("#27ae60", "#f39c12", "#c0392b"),
      domain  = c(0, 4)
    ) |>
    data_color(
      columns = mean_influence,
      method  = "numeric",
      palette = c("#27ae60", "#f39c12", "#c0392b"),
      domain  = c(0, 1)
    )
}
```

```{r innovation-time-plot}
if (!is.null(inn)) {
  inn |>
    mutate(depth_label = if_else(is.na(depth), variable,
                                  paste0(variable, " (", depth, " m)"))) |>
    ggplot(aes(x = time, y = innovation, colour = variable)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_point(alpha = 0.5, size = 1.2) +
    geom_smooth(method = "loess", span = 0.4, se = FALSE, linewidth = 0.8) +
    facet_wrap(~variable, scales = "free_y") +
    labs(title = "Innovation time series by observation type",
         y = "Innovation (obs − predicted)", x = NULL) +
    theme(legend.position = "none")
}
```

```{r nis-single-plot}
if (!is.null(inn)) {
  inn |>
    ggplot(aes(x = variable, y = nis_single, fill = variable)) +
    geom_violin(alpha = 0.5, draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
    geom_hline(yintercept = 2, linetype = "dotted", colour = "#d9a520") +
    scale_y_log10(limits = c(0.01, NA)) +
    labs(title = "Per-observation NIS by obs type  (ideal = 1)",
         y = "NIS (log scale)", x = NULL) +
    theme(legend.position = "none")
}
```

### Observation influence by depth

> Influence close to 1 means the observation is nearly fully controlling its
> own state update — there is little model contribution. Values above ~0.8 suggest
> `obs_sd` may be too small for that variable/depth.

```{r obs-influence-plot}
if (!is.null(inn) && any(!is.na(inn$depth))) {
  inn |>
    filter(!is.na(depth)) |>
    group_by(variable, depth) |>
    summarise(mean_influence = mean(obs_influence, na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = depth, y = mean_influence, colour = variable)) +
    geom_point(size = 3) +
    geom_line() +
    geom_hline(yintercept = 0.8, linetype = "dashed", colour = "#c0392b", linewidth = 0.6) +
    coord_flip() +
    scale_x_reverse() +
    labs(title = "Mean observation influence by variable and depth",
         subtitle = "Dashed line = 0.8 (high influence threshold)",
         x = "Depth (m)", y = "Mean obs influence (diag of HK)") +
    facet_wrap(~variable, nrow = 1)
}
```

---

## 3 Ensemble Spread {#sec-spread}

> The **spread ratio** (posterior SD / prior SD) should be less than 1 (the filter
> reduces uncertainty). Ratios near 0 indicate ensemble collapse — the filter is
> becoming overconfident and will not respond well to future observations.

```{r spread-ratio-heatmap}
if (!is.null(spr)) {
  spr_summary <- spr |>
    group_by(variable, depth) |>
    summarise(
      mean_spread_ratio = mean(spread_ratio, na.rm = TRUE),
      mean_prior_sd     = mean(prior_sd,     na.rm = TRUE),
      .groups = "drop"
    )

  top_states <- spr_summary |>
    group_by(variable) |>
    summarise(mean_spread_ratio = mean(mean_spread_ratio), .groups = "drop") |>
    arrange(mean_spread_ratio) |>
    slice_head(n = 12) |>
    pull(variable)

  spr_summary |>
    filter(variable %in% top_states) |>
    ggplot(aes(x = depth, y = reorder(variable, mean_spread_ratio),
               fill = mean_spread_ratio)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    scale_fill_gradient2(
      low = "#c0392b", mid = "#f39c12", high = "#27ae60",
      midpoint = 0.5, limits = c(0, 1),
      name = "Spread\nratio"
    ) +
    labs(title = "Mean spread ratio (posterior SD / prior SD) by state × depth",
         subtitle = "Red = collapse risk  |  Green = healthy reduction",
         x = "Depth (m)", y = NULL) +
    theme(legend.position = "right")
}
```

```{r prior-sd-collapse}
if (!is.null(spr)) {
  collapse_states <- spr |>
    filter(spread_ratio < 0.05) |>
    distinct(variable) |>
    pull(variable)

  if (length(collapse_states) > 0) {
    spr |>
      filter(variable %in% collapse_states) |>
      group_by(time, variable) |>
      summarise(mean_prior_sd = mean(prior_sd, na.rm = TRUE), .groups = "drop") |>
      ggplot(aes(x = time, y = mean_prior_sd, colour = variable)) +
      geom_line() +
      labs(title = "Prior SD over time for states with spread_ratio < 0.05  (collapse risk)",
           y = "Mean prior SD across depths", x = NULL)
  }
}
```

```{r update-magnitude-plot}
if (!is.null(spr)) {
  spr |>
    group_by(time, variable) |>
    summarise(mean_update = mean(update_magnitude, na.rm = TRUE), .groups = "drop") |>
    filter(variable %in% head(
      spr |> group_by(variable) |>
        summarise(m = mean(update_magnitude, na.rm = TRUE), .groups = "drop") |>
        arrange(desc(m)) |> pull(variable),
      8
    )) |>
    ggplot(aes(x = time, y = mean_update, colour = variable)) +
    geom_line(alpha = 0.8) +
    facet_wrap(~variable, scales = "free_y", ncol = 2) +
    labs(title = "Mean update magnitude over time (top 8 states)",
         y = "| posterior_mean − prior_mean |", x = NULL) +
    theme(legend.position = "none")
}
```

---

## 4 Rank Histograms {#sec-ranks}

> A rank histogram (Talagrand diagram) tests ensemble calibration. If the filter
> is well-calibrated, the observation rank is **uniformly distributed** over
> 1 : (N+1). A **peaked** histogram (too many middle ranks) means the ensemble
> is under-dispersive. A **U-shaped** histogram means it is over-dispersive.

```{r rank-histograms}
if (!is.null(rnk)) {
  # Infer ensemble size from maximum rank observed
  n_ens_inferred <- max(rnk$obs_rank, na.rm = TRUE) - 1L
  n_bins <- min(n_ens_inferred + 1L, 20L)

  rank_chi_sq <- rnk |>
    group_by(variable) |>
    summarise(
      n = n(),
      chi_p = tryCatch(
        chisq.test(table(cut(obs_rank,
                             breaks = seq(0.5, n_ens_inferred + 1.5, length.out = n_bins + 1),
                             labels = FALSE)))$p.value,
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) |>
    mutate(calibration = case_when(
      is.na(chi_p)  ~ "insufficient data",
      chi_p > 0.10  ~ "uniform (well-calibrated)",
      chi_p > 0.01  ~ "marginal (p < 0.10)",
      TRUE          ~ "non-uniform (p < 0.01)"
    ))

  rnk |>
    ggplot(aes(x = obs_rank)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = n_bins, fill = "#2980b9", colour = "white", alpha = 0.8
    ) +
    geom_hline(
      aes(yintercept = 1 / (n_ens_inferred + 1)),
      linetype = "dashed", colour = "#c0392b", linewidth = 0.7
    ) +
    facet_wrap(~variable, scales = "free_y") +
    labs(
      title = sprintf("Rank histograms  (ensemble size ≈ %d)", n_ens_inferred),
      subtitle = "Dashed line = expected uniform density  |  Peaked = under-spread  |  U-shaped = over-spread",
      x = "Rank of observation within prior ensemble",
      y = "Density"
    )
}
```

```{r rank-calibration-table}
if (!is.null(rnk)) {
  rank_chi_sq |>
    gt() |>
    tab_header(title = "Rank Histogram Calibration Test (chi-squared uniformity)") |>
    cols_label(variable = "Obs type", n = "N", chi_p = "p-value",
               calibration = "Status") |>
    data_color(
      columns = calibration,
      method  = "factor",
      palette = c("#c0392b", "#f39c12", "#27ae60", "#bdc3c7")
    )
}
```

---

## 5 Parameter Estimation {#sec-parameters}

```{r params-section}
#| results: asis
if (!has_pars) {
  cat("::: {.callout-note}\nNo parameters were estimated in this run ",
      "(or `da_parameters.parquet` was not found).\n:::\n")
}
```

```{r param-evolution}
if (has_pars) {
  par |>
    ggplot(aes(x = time)) +
    geom_ribbon(aes(ymin = prior_mean - prior_sd,
                    ymax = prior_mean + prior_sd),
                fill = "#3498db", alpha = 0.2) +
    geom_line(aes(y = prior_mean), colour = "#3498db", linewidth = 0.8) +
    geom_ribbon(aes(ymin = posterior_mean - posterior_sd,
                    ymax = posterior_mean + posterior_sd),
                fill = "#e74c3c", alpha = 0.2) +
    geom_line(aes(y = posterior_mean), colour = "#e74c3c", linewidth = 0.8) +
    facet_wrap(~parameter, scales = "free_y", ncol = 2) +
    labs(title = "Parameter evolution  (blue = prior, red = posterior, ribbons = ±1 SD)",
         y = "Value", x = NULL)
}
```

```{r param-bounds}
if (has_pars && any(par$frac_at_lower_bound > 0 | par$frac_at_upper_bound > 0,
                    na.rm = TRUE)) {
  par |>
    select(time, parameter, frac_at_lower_bound, frac_at_upper_bound) |>
    pivot_longer(c(frac_at_lower_bound, frac_at_upper_bound),
                 names_to = "bound", values_to = "frac") |>
    mutate(bound = recode(bound,
                          frac_at_lower_bound = "lower",
                          frac_at_upper_bound = "upper")) |>
    filter(frac > 0) |>
    ggplot(aes(x = time, y = frac, colour = bound)) +
    geom_line() +
    geom_point(size = 1.5) +
    geom_hline(yintercept = 0.05, linetype = "dashed", colour = "#c0392b") +
    facet_wrap(~parameter, ncol = 2) +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Fraction of ensemble members at parameter bounds",
         subtitle = "Dashed = 5% threshold; above this, widen bounds or change par_fit_method",
         y = "Fraction at bound", x = NULL)
}
```

```{r param-obs-contribution}
if (has_pars && "phy_contribution" %in% names(par)) {
  par |>
    select(time, parameter, phy_contribution, other_contribution) |>
    pivot_longer(c(phy_contribution, other_contribution),
                 names_to = "source", values_to = "contribution") |>
    mutate(source = recode(source,
                           phy_contribution   = "PHY_TCHLA obs",
                           other_contribution = "Other obs")) |>
    ggplot(aes(x = time, y = contribution, fill = source)) +
    geom_col(position = "stack") +
    geom_hline(yintercept = 0, colour = "grey30") +
    facet_wrap(~parameter, scales = "free_y", ncol = 2) +
    labs(title = "Parameter update decomposition: PHY_TCHLA vs other observations",
         subtitle = "Dominant 'Other obs' for a PHY parameter suggests spurious cross-variable updating",
         y = "Mean update contribution", x = NULL) +
    scale_fill_manual(values = c("PHY_TCHLA obs" = "#27ae60", "Other obs" = "#e74c3c"),
                      name = NULL)
}
```

---

## 6 Cross-State & Cross-Depth Coupling {#sec-coupling}

> These diagnostics are only available when **PHY_TCHLA** (composite chlorophyll)
> observations are active. They reveal how phytoplankton observations propagate to
> nutrient/oxygen states (**cross-state**) and how observations at one depth update
> states at other depths (**cross-depth**).

```{r coupling-section}
#| results: asis
if (!has_cross) {
  cat("::: {.callout-note}\n")
  cat("No cross-state or cross-depth data found. This section requires ",
      "PHY_TCHLA (composite chlorophyll) observations to be active.\n")
  cat(":::\n")
}
```

```{r cross-state-plot}
if (has_cross) {
  # Derive state group (NIT, PHS, OXY, etc.) for colour faceting
  cst_plot <- cst |>
    mutate(
      state_group = sub("_.*", "", variable),
      sign_label  = case_when(
        is.na(sign_consistent) ~ "near-zero",
        sign_consistent        ~ "consistent",
        TRUE                   ~ "inconsistent"
      )
    )

  ggplot(cst_plot, aes(x = time, y = cov_with_phy_hx, colour = state_group)) +
    geom_hline(yintercept = 0, colour = "grey50", linetype = "dashed") +
    geom_line(linewidth = 0.6, alpha = 0.7) +
    geom_point(aes(shape = sign_label), size = 1.8) +
    facet_wrap(~variable, scales = "free_y", ncol = 3) +
    scale_shape_manual(
      values = c(consistent = 16, "near-zero" = 1, inconsistent = 4),
      name   = "K sign vs Cov"
    ) +
    labs(
      title    = "Cross-state covariance: Cov(state, PHY_TCHLA H×x)",
      subtitle = paste0(
        "Cov < 0: drawdown phase (high PHY, low nutrients)   ",
        "Cov > 0: growth phase (high nuts fuel high PHY)\n",
        "Shape: whether Kalman gain sign matches covariance sign"
      ),
      y = "Covariance", x = NULL, colour = "State group"
    )
}
```

```{r cross-state-update}
if (has_cross) {
  cst |>
    ggplot(aes(x = time, y = mean_update, fill = sub("_.*", "", variable))) +
    geom_col(position = "dodge") +
    geom_hline(yintercept = 0, colour = "grey30") +
    facet_wrap(~variable, scales = "free_y", ncol = 3) +
    labs(
      title    = "Mean update applied to coupled states via PHY_TCHLA observations",
      subtitle = "Non-zero updates confirm cross-variable information propagation",
      y = "Mean update (K × innovation)", x = NULL,
      fill = "State group"
    )
}
```

```{r cross-depth-heatmap}
if (!is.null(cdp) && nrow(cdp) > 0) {
  # Average update contribution over all DA steps for each obs×state depth pair
  cdp_avg <- cdp |>
    group_by(phy_state, obs_depth, state_depth) |>
    summarise(mean_contrib = mean(update_contribution, na.rm = TRUE),
              .groups = "drop") |>
    mutate(
      obs_label   = paste0("obs ", round(obs_depth, 1), " m"),
      state_label = paste0(round(state_depth, 1), " m")
    )

  ggplot(cdp_avg,
         aes(x = state_label, y = obs_label, fill = mean_contrib)) +
    geom_tile(colour = "white", linewidth = 0.3) +
    geom_text(aes(label = sprintf("%+.3f", mean_contrib)),
              size = 2.5, colour = "grey20") +
    facet_wrap(~phy_state, ncol = 2) +
    scale_fill_gradient2(
      low = "#2166ac", mid = "white", high = "#d6604d",
      midpoint = 0, name = "Mean\nK×innov"
    ) +
    labs(
      title    = "Cross-depth update structure (averaged over all DA steps)",
      subtitle = paste0(
        "rows = PHY_TCHLA obs depth   cols = state depth   ",
        "cell = K[state_d, obs_d] × innov\n",
        "Same-depth diagonal should dominate; off-diagonal = vertical spreading"
      ),
      x = "State depth", y = "Obs depth"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid  = element_blank())
}
```

---

## 7 Summary and Recommendations {#sec-recommendations}

```{r recommendation-engine}
#| results: asis

findings <- list()

# ── NIS ──────────────────────────────────────────────────────────────────────
if (!is.null(fh)) {
  mean_nis   <- mean(fh$nis, na.rm = TRUE)
  pct_hi_nis <- mean(fh$nis > 2,   na.rm = TRUE)
  pct_lo_nis <- mean(fh$nis < 0.5, na.rm = TRUE)

  if (mean_nis > 2 || pct_hi_nis > 0.3) {
    findings[["nis_high"]] <- list(
      severity = "critical",
      symptom  = sprintf("NIS consistently too large (mean = %.2f; %.0f%% of steps > 2)",
                         mean_nis, pct_hi_nis * 100),
      meaning  = "The ensemble spread is **too narrow** — the filter is overconfident. Innovations are much larger than the ensemble predicts.",
      actions  = c(
        "**Increase `model_sd`** in `depth_model_sd.csv` or `states_config.csv` (process noise) — try multiplying current values by 1.5×.",
        "**Increase `inflation_factor`** in `configure_flare.yml` (e.g., from 1.0 to 1.05–1.10).",
        "**Increase `ensemble_size`** — small ensembles underestimate covariance.",
        "**Increase `vert_decorr_length`** if NIS is elevated only at deeper layers, suggesting noise isn't propagating vertically."
      )
    )
  } else if (mean_nis < 0.5 || pct_lo_nis > 0.3) {
    findings[["nis_low"]] <- list(
      severity = "warning",
      symptom  = sprintf("NIS consistently too small (mean = %.2f; %.0f%% of steps < 0.5)",
                         mean_nis, pct_lo_nis * 100),
      meaning  = "The ensemble spread is **too large** — the filter is over-inflated. Observations are well within the ensemble range and barely move the analysis.",
      actions  = c(
        "**Reduce `inflation_factor`** (e.g., from 1.10 to 1.02–1.05) or disable inflation entirely.",
        "**Reduce `model_sd`** — the process noise may be adding too much spread each step.",
        "**Check `obs_sd`** in `observations_config.csv` — inflated obs error allows innovations to stay small without any model adjustment needed."
      )
    )
  }
}

# ── Ensemble spread / collapse ────────────────────────────────────────────────
if (!is.null(spr)) {
  collapsed <- spr |>
    group_by(variable) |>
    summarise(min_ratio = min(spread_ratio, na.rm = TRUE),
              mean_ratio = mean(spread_ratio, na.rm = TRUE),
              .groups = "drop") |>
    filter(mean_ratio < 0.05)

  if (nrow(collapsed) > 0) {
    findings[["collapse"]] <- list(
      severity = "critical",
      symptom  = paste0("Ensemble collapse detected for: ",
                        paste(collapsed$variable, collapse = ", ")),
      meaning  = "Spread ratio near zero means the ensemble has become degenerate — all members agree and the filter **cannot assimilate new observations** for these states.",
      actions  = c(
        "**Increase `model_sd`** for the collapsed states in `states_config.csv`.",
        "**Increase `ensemble_size`** — collapse is more likely with small ensembles.",
        "**Check `par_fit_method`**: if set to `perturb_const`, switching to `perturb` may help.",
        "**Widen parameter bounds** if parameter collapse is driving state collapse."
      )
    )
  }
}

# ── Rank histogram calibration ────────────────────────────────────────────────
if (!is.null(rnk)) {
  n_ens_inf <- max(rnk$obs_rank, na.rm = TRUE) - 1L
  n_bins_r  <- min(n_ens_inf + 1L, 20L)

  rank_stats <- rnk |>
    group_by(variable) |>
    summarise(
      n           = n(),
      chi_p       = tryCatch(
        chisq.test(table(cut(obs_rank,
                             breaks = seq(0.5, n_ens_inf + 1.5, length.out = n_bins_r + 1),
                             labels = FALSE)))$p.value,
        error = function(e) NA_real_
      ),
      # peaked = high central mass = under-dispersive
      central_frac = mean(obs_rank > n_ens_inf * 0.25 & obs_rank < n_ens_inf * 0.75,
                          na.rm = TRUE),
      # U-shaped = high tail mass = over-dispersive
      tail_frac    = mean(obs_rank <= max(1, n_ens_inf * 0.1) |
                          obs_rank >= n_ens_inf * 0.9,
                          na.rm = TRUE),
      .groups = "drop"
    ) |>
    filter(!is.na(chi_p), chi_p < 0.05)

  peaked_vars <- rank_stats |>
    filter(central_frac > 0.6) |>
    pull(variable)

  ushaped_vars <- rank_stats |>
    filter(tail_frac > 0.3) |>
    pull(variable)

  if (length(peaked_vars) > 0) {
    findings[["rank_peaked"]] <- list(
      severity = "warning",
      symptom  = paste0("Non-uniform (peaked) rank histogram for: ",
                        paste(peaked_vars, collapse = ", ")),
      meaning  = "The ensemble is **under-dispersive** for these obs types. The observation is consistently in the middle of the ensemble — it is never surprising.",
      actions  = c(
        "**Increase `model_sd`** for the corresponding states.",
        "**Decrease `obs_sd`** slightly — if obs error is overestimated, the perturbed observations cluster near the truth and the ensemble fills out too broadly.",
        "**Check `vert_decorr_length`** — too-short vertical correlation adds noise without useful structure."
      )
    )
  }

  if (length(ushaped_vars) > 0) {
    findings[["rank_ushaped"]] <- list(
      severity = "warning",
      symptom  = paste0("Non-uniform (U-shaped) rank histogram for: ",
                        paste(ushaped_vars, collapse = ", ")),
      meaning  = "The ensemble is **over-dispersive** for these obs types. The observation frequently falls outside the ensemble — the ensemble is too wide.",
      actions  = c(
        "**Reduce `model_sd`** for the corresponding states.",
        "**Reduce `inflation_factor`** or disable inflation.",
        "**Increase `obs_sd`** if observation errors may be underestimated."
      )
    )
  }
}

# ── Observation influence ─────────────────────────────────────────────────────
if (!is.null(inn)) {
  high_inf <- inn |>
    group_by(variable) |>
    summarise(mean_inf = mean(obs_influence, na.rm = TRUE), .groups = "drop") |>
    filter(mean_inf > 0.8)

  if (nrow(high_inf) > 0) {
    findings[["high_influence"]] <- list(
      severity = "warning",
      symptom  = paste0("Very high observation influence (>0.8) for: ",
                        paste(high_inf$variable, collapse = ", ")),
      meaning  = "These observations are nearly fully controlling the analysis — the model ensemble is contributing almost nothing. This risks filter divergence if observation error is underspecified.",
      actions  = c(
        "**Increase `obs_sd`** for the high-influence variables in `observations_config.csv`.",
        "**Increase ensemble spread** (via `model_sd` or `inflation_factor`) so the prior is wider.",
        "**Enable localization** (`localization_distance` in `configure_flare.yml`) to prevent a single obs from updating distant depths."
      )
    )
  }
}

# ── Parameter bounds ─────────────────────────────────────────────────────────
if (has_pars) {
  bound_issues <- par |>
    group_by(parameter) |>
    summarise(
      max_lower = max(frac_at_lower_bound, na.rm = TRUE),
      max_upper = max(frac_at_upper_bound, na.rm = TRUE),
      .groups = "drop"
    ) |>
    filter(max_lower > 0.05 | max_upper > 0.05)

  if (nrow(bound_issues) > 0) {
    findings[["par_bounds"]] <- list(
      severity = "warning",
      symptom  = paste0("Parameter ensemble hitting bounds (>5% of members) for: ",
                        paste(bound_issues$parameter, collapse = ", ")),
      meaning  = "Members piling up at bounds reduce ensemble spread and can cause filter collapse for that parameter.",
      actions  = c(
        "**Widen bounds** in `parameter_calibration_config.csv`.",
        "**Switch `par_fit_method`** from `perturb` to `perturb_const` to stabilise perturbation magnitude.",
        "**Check spurious cross-variable updating** — if 'Other obs' dominates in the update decomposition, the parameter may be driven by biologically unrelated observations."
      )
    )
  }
}

# ── Render findings ───────────────────────────────────────────────────────────
if (length(findings) == 0) {
  cat("::: {.callout-tip icon=true}\n")
  cat("## Filter appears healthy\n")
  cat("No significant issues detected across NIS, ensemble spread, rank histograms, ")
  cat("observation influence, or parameter bounds.\n")
  cat(":::\n\n")
} else {
  severity_icon <- c(critical = "🔴", warning = "🟡", info = "🔵")
  severity_type <- c(critical = "callout-important", warning = "callout-warning",
                     info = "callout-note")

  for (key in names(findings)) {
    f   <- findings[[key]]
    sev <- f$severity
    cat(sprintf("::: {.%s}\n", severity_type[sev]))
    cat(sprintf("## %s %s\n\n", severity_icon[sev], f$symptom))
    cat("**What this means:** ", f$meaning, "\n\n", sep = "")
    cat("**Recommended actions:**\n\n")
    for (a in f$actions) cat(sprintf("- %s\n", a))
    cat(":::\n\n")
  }
}
```

---

## 8 Hyperparameter Reference {#sec-reference}

```{r hyperparameter-table}
tibble::tribble(
  ~Parameter,           ~`Config file`,                     ~`Typical range`,   ~`Too small → symptom`,                    ~`Too large → symptom`,
  "ensemble_size",      "configure_flare.yml",              "31–500",           "Noisy covariance; needs inflation",        "Slow runtime",
  "model_sd",           "states_config.csv / depth_model_sd.csv", "0.1–2.0 (var-dependent)", "Collapse; NIS >> 1; peaked rank hist", "Over-spread; NIS << 1; U-shaped rank hist",
  "obs_sd",             "observations_config.csv",          "0.3–1.0 (var-dependent)", "High influence; filter divergence risk", "DA barely moves the ensemble",
  "inflation_factor",   "configure_flare.yml",              "1.0–1.15",         "Ensemble collapse",                        "Over-spread; NIS << 1",
  "localization_distance", "configure_flare.yml",           "2–10 m",           "Spurious long-range updates (disabled = ∞)", "Information doesn't propagate vertically",
  "vert_decorr_length", "states_config.csv",                "1–10 m",           "Layer-to-layer noise; non-physical",       "Over-smooth vertical profiles",
  "par_lowerbound / par_upperbound", "parameter_calibration_config.csv", "domain-specific", "Members hit wall; collapse",     "Implausible values enter ensemble"
) |>
  gt() |>
  tab_header(
    title    = "Hyperparameter Quick Reference",
    subtitle = "FLARE EnKF filter tuning guide"
  ) |>
  tab_style(
    style    = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style    = cell_fill(color = "#fef3cd"),
    locations = cells_body(columns = `Too small → symptom`)
  ) |>
  tab_style(
    style    = cell_fill(color = "#d1ecf1"),
    locations = cells_body(columns = `Too large → symptom`)
  ) |>
  cols_width(Parameter ~ px(200), `Config file` ~ px(220))
```

---

*Generated by `FLAREr::write_da_diagnostics()`. Enable with `save_da_diagnostics: TRUE` in `configure_flare.yml`.*
